## Functions and Conditionals

### Conditionals

The basic syntax for an if statement goes like this:

```{r}
if (logical statement) {
  do something
} else {
  do something else
}
```

The logical statements should boil down to a boolean; for example:

```{r}
a <- 10
if (a <= 5) {
  print(a)
} else {
  print ("It's more than 5.")
}
```

We can also add else if statements in the middle if we need more options.

```{r}
a <- 15

if (a <= 5) {
  print("small")
} else if (a > 5 & a <= 10) {
  print("medium")
} else if (a > 10 | a == 7) {
  print ("large")
} else {
  print("huge")
}
```

Neat! We can also nest conditionals just like we do in Python.

```{r}
a <- 5
if (a/2 != 0) {
  if (a ==5){
    print("a is 5.")
  }
}
```

Neat! Now, let's look at another core part of R programming - Loops!

### Loops

```{r}
for (value in 1:5) {
  # Paste is a concatenation function.
  print(paste(value, "is greater than", value - 1))
}
```

We can also do nifty stuff with our vectors, such as apply changes to specific indices of our data.

```{r}
a_vector <- c(1,1,1,1,1,1,1)

# Change the first, third, and sixth elements to the value 6.

for (i in c(1,3,6)){
  a_vector[i] <- 6
}

a_vector
```

We can loop over more than just lists:

```{r}
mat <- matrix(1, nrow = 2, ncol = 2)

for (row in 1:2){
  for(col in 1:2){
    mat[row, col] <- 0
    print(mat)
  }
}
```

Like Python, we have while loops in R:

```{r}
b <- 0
while (b < 10) {
  I # Increase b by 2.
  b <- b + 2
  # Print b.
  print(b)
}
```

## Functions

When should we use functions?

```{r}
our_function <- function(optional_params){
  # Do some stuff.
}
```

An example:

```{r}
test_temp <- c(16.550540, 26.942005, 32.591088, 8.580116, 28.003873, 28.542391, 21.173577, 21.048836, 18.769735)

# Then we want a function that converts from C to F:
c_to_f <- function(dat){
  temp_fahr <- dat*(9/5) + 32
  
  return(temp_fahr)
}

print(c_to_f(26))
```

This is all cool.. but it didn't do anything. We need to call the function!

```{r}
data_fahr <- c_to_f(test_temp)
data_fahr
```

Notice that in the background here, we are looping over each item in the array, but we never write a for loop! Keep this in mind, since when you pass in a large data set, the changes made by functions aren't quite immediate - the computer does have to go over each item in your data set, and change them one at a time.

A shortcut of sorts is to use the `apply()` function, which operates very similarly to Python's `map()`. Note that `apply` syntax goes as follows: `apply(X, MARGIN, FUN)` where X is the data set to use, MARGIN is specification for rows or columns, and FUN is the function to apply.

```{r}
test_mat <- matrix(c(1,1,1,2,2,2), nrow <- 2)
test_mat
# 1 is for rows, 2 is for columns.
apply(test_mat, 2, sum)
```

It is common to use `mean`, `median`, `sum`, `min`, or `max` with `apply`, but we can also use our user-defined functions as well. Note that apply() only works with matrices.

```{r}
test_temp_mat <- matrix(c(16.550540, 26.942005, 32.591088, 8.580116, 28.003873, 28.542391, 20.976820, 21.173577, 21.048836, 18.769735), nrow = 2)

apply(test_temp_mat, 1, c_to_f)
```

### Homework

1.  Do 5 CodeWars problems in R. Share your solutions here:

    ```{r}

    # 1. Create a function that takes an integer as an argument and returns "Even" for even numbers or "Odd" for odd numbers.

    even_or_odd = function(number) {
      if (number %% 2 == 0) {
        print("Even")
    } else {
        print("Odd")
    }
    }

    # 2. In this little assignment you are given a string of space separated numbers, and have to return the highest and lowest number.

    # Examples:
    # Input: "1 2 3 4 5"   =>  Output: "5 1"
    # Input: "1 2 -3 4 5"  =>  Output: "5 -3"
    # Input: "1 9 3 4 -5"  =>  Output: "9 -5"

    high_and_low <- function(numbers){
      number_strings <- strsplit(numbers, " ")[[1]]
      numbers <- as.numeric(number_strings)
      max_number <- max(numbers, na.rm = TRUE)
      min_number <- min(numbers, na.rm = TRUE)
      return(paste(max_number, min_number))
    }

    # 3. Implement a function that accepts 3 integer values a, b, c. The function should return true if a triangle can be built with the sides of given length and false in any other case.

    # (In this case, all triangles must have surface greater than 0 to be accepted).

    # Examples:
    # Input -> Output
    # 1,2,2 -> true
    # 4,2,3 -> true
    # 2,2,2 -> true
    # 1,2,3 -> false
    # -5,1,3 -> false
    # 0,2,3 -> false
    # 1,2,9 -> false

    is_triangle <- function(a, b, c){
      if (a <= 0 || b <= 0 || c <= 0) {
        return(FALSE)
      }
      if ((a + b > c) && (a + c > b) && (b + c > a)) {
        return(TRUE)
      }
      else {
        return(FALSE)
      }
    }

    # 4. Complete the solution so that it splits the string into pairs of two characters. If the string contains an odd number of characters then it should replace the missing second character of the final pair with an underscore ('_').

    # Examples:
    # 'abc' =>  ['ab', 'c_']
    # 'abcdef' => ['ab', 'cd', 'ef']

    solution <- function(s) {
      result <- c()
      len <- nchar(s)
      if (len == 0) {
        return(result)
      }
      for (i in seq(1, len, by = 2)) {
        pair <- substring(s, i, min(i + 1, len))
        if (nchar(pair) == 1) {
          pair <- paste0(pair, "_")
        }
        result <- c(result, pair)
      }
      return(result)
    }

    # 5. Given three integers a, b, and c, return the largest number obtained after inserting the operators +, *, and parentheses (). In other words, try every combination of a, b, and c with the operators, without reordering the operands, and return the maximum value.

    # Example
    # With the numbers 1, 2, and 3, here are some possible expressions:
    # 1 * (2 + 3) = 5
    # 1 * 2 * 3 = 6
    # 1 + 2 * 3 = 7
    # (1 + 2) * 3 = 9
    # The maximum value that can be obtained is 9.

    expression_matter <- function(a, b, c) {
      result1 <- a * (b + c)
      result2 <- a * b * c
      result3 <- a + b * c
      result4 <- (a + b) * c
      result5 <- a + b + c
      result6 <- a * (b + c)
      
      max_result <- max(result1, result2, result3, result4, result5, result6)
      
      return(max_result)
    }
    ```

2.  Create a function that checks an input of type `int` and returns fizz if the number is a multiple of 3, buzz if the number is a multiple of 5, and fizzbuzz if the number is a multiple of 3 and 5.

    ```{r}

    fizzbuzz_check <- function(input) {
      if (input %% 3 == 0 & input %% 5 == 0) {
        print("FIZZBUZZ")
    } else if (input %% 5 == 0) {
        print("BUZZ")
    } else if (input %% 3 == 0) {
        print ("FIZZ")
    } else {
        print("Your input is not divisible by 3 or 5.")
    }
    }
    ```

3.  **Mario Challenge:** In the classic Super Mario games, Mario ends each level by ascending a flight of stairs and leaping off the top to arrive at the castle at the end. (See [here](https://youtu.be/-avspZlbOWU?t=55) for details). We can imagine that each of the bricks of the stairs is a character X. Create a function that prints out a set of stairs. For example:

    If the function is called marioStairs() and we pass in the value 4, we would expect to see stairs with a length of 4 and a height of 4:

    X

    XX

    XXX

    XXXX

    ```{r}

    # You may feel more comfortable figuring this out in Python first, then translating into R later.

    marioStairs <- function(stairs) {
      for (i in 1:stairs) {
        cat(rep("X", i), "\n")
      }
    }

    ```
